---
title: "Isotope analytical uncertainty propagation"
author: "Ruan van Mazijk"
output:
  pdf_document:
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```

```{r}
source(here::here("analyses/mass-balance.R"))
```

Here I (1) state the uncertainty propagation rules plainly, (2) apply them to our mass-balance model equation and calculate the results, and finally (3) I explore the the derivation of Genereux's (1998) isotopic tracer component uncertainty equations and how they, sadly, differ to my arithmetic propagation in section (2).

# Uncertainty propagation rules

Our uncertainty rules are as follows. Let $q$ be some quantity, and $u_q$ be its uncertainty, where $q$ is a function of variables $x$, $y$ and $z$.

For **constants**

\begin{align*} \tag{1.1} \label{1.1}
  q &= Bx \\
  \frac{u_q}{q} &= \frac{u_x}{x}
\end{align*}

For **sums**

\begin{align*} \tag{1.2} \label{1.2}
  q &= x \pm y \pm z \\
  u_q &= \sqrt{u_x^2 +u_y^2 + u_z^2}
\end{align*}

And for **products/quotients**

\begin{align*} \tag{1.3} \label{1.3}
  q &= xyz \\
  \frac{u_q}{|q|} &= \sqrt{
    \left( \frac{u_x}{x} \right)^2 +
    \left( \frac{u_y}{y} \right)^2 +
    \left( \frac{u_z}{z} \right)^2
  }
\end{align*}

\break

# Ruan's propagation

## Derivation {-}

Let $p_E$ be the proportion of streamflow derived from rainfall according to an isotope $E$, where

\begin{align*} \tag{2.1} \label{2.1}
  p_E = \frac{E_{streamflow} - E_{baseflow}}
                {E_{rain} - E_{baseflow}}
\end{align*}

As such, for our analysis, we propagate the uncertainty as follows, relying on our known uncertainty in measuring isotope values $u_E$

\begin{align*}
  p_E &= \frac{E_{streamflow} - E_{baseflow}}
              {E_{rain} - E_{baseflow}} \\
  \therefore u_{p_E} &= |p_E| \sqrt{
    \left( \frac{u_{[E_{streamflow} - E_{baseflow}]}}
                {E_{streamflow} - E_{baseflow}} \right)^2 +
    \left( \frac{u_{[E_{rain} - E_{baseflow}]}}
                {E_{rain} - E_{baseflow}} \right)^2
  }
\end{align*}

where

\begin{align*}
  u_{[E_{streamflow} - E_{baseflow}]} = u_{[E_{rain} - E_{baseflow}]} = \sqrt{u_E^2 + u_E^2}
\end{align*}

such that

\begin{align*} \tag{2.2} \label{2.2}
  u_{p_E} = |p_E| \sqrt{
    \left( \frac{\sqrt{u_E^2 + u_E^2}}
                {E_{streamflow} - E_{baseflow}} \right)^2 +
    \left( \frac{\sqrt{u_E^2 + u_E^2}}
                {E_{rain} - E_{baseflow}} \right)^2
  }
\end{align*}

We then average $p_{\delta^{2}H}$ and $p_{\delta^{18}O}$, each with their own uncertainty derived above, as follows

\begin{align*} \tag{2.3} \label{2.3}
  p &= \frac{p_{\delta^{18}O} + p_{\delta^{2}H}}
            {2} \\
  \therefore
  u_p &= |p| \frac{\sqrt{u_{p_{\delta^{18}O}} + u_{p_{\delta^{2}H}}}}
                  {p_{\delta^{18}O} + p_{\delta^{2}H}}
\end{align*}

\break

## Application to Liesbeek study {-}

For our study, we combined long term analytical precision and accuracy using Equation \ref{1.2}

\begin{align*}
  u_{\delta^{18}O} &= \sqrt{`r precision$d18O`^2 + `r accuracy$d18O`^2} \\
  &= `r u$d18O` \\
  u_{\delta^{2}H} &= \sqrt{`r precision$d2H`^2 + `r accuracy$d2H`^2} \\
  &= `r u$d2H`
\end{align*}

As such, we calculate our $p_E$ values using Equation \ref{2.1} as

\begin{align*}
  p_{\delta^{18}O} &= \frac{`r streamflow$d18O` - `r baseflow$d18O`}
                           {`r rain$d18O` - `r baseflow$d18O`} \\
  &= `r p$d18O` \\
  p_{\delta^{2}H} &= \frac{`r streamflow$d2H` - `r baseflow$d2H`}
                          {`r rain$d2H` - `r baseflow$d2H`} \\
  &= `r p$d2H`
\end{align*}

and their uncertainties using Equation \ref{2.2}

\begin{align*}
  u_{p_{\delta^{18}O}} &= |`r p$d18O`| \sqrt{
    \left( \frac{\sqrt{`r u$d18O`^2 + `r u$d18O`^2}}
                {`r streamflow$d18O` - `r baseflow$d18O`} \right) ^ 2 +
    \left( \frac{\sqrt{`r u$d18O`^2 + `r u$d18O`^2}}
                {`r rain$d18O` - `r baseflow$d18O`} \right) ^ 2
  } \\
  &= `r u_p$d18O` \\
  u_{p_{\delta^{2}H}} &= |`r p$d2H`| \sqrt{
    \left( \frac{\sqrt{`r u$d2H`^2 + `r u$d2H`^2}}
                {`r streamflow$d2H` - `r baseflow$d2H`} \right) ^ 2 +
    \left( \frac{\sqrt{`r u$d2H`^2 + `r u$d2H`^2}}
                {`r rain$d2H` - `r baseflow$d2H`} \right) ^ 2
  } \\
  &= `r u_p$d2H`
\end{align*}

We then use Equation \ref{2.3} to derive $p$ proper

\begin{align*}
  p &= \frac{`r p$d18O` + `r p$d2H`}
            {2} \\
  &= `r p_rain` \\
  \therefore 
  u_p &= |`r p_rain`| \frac{\sqrt{`r u_p$d18O` + `r u_p$d2H`}}
                           {`r p$d18O` + `r p$d2H`} \\
  &= `r u_p_rain`
\end{align*}

Thus, we can conclude that, following this method, the Liesbeek River streamflow during our storm constituted 101% ± 12.3% rain-water.

\break

# Genereux's propagation

Knowing Equations 1--3, applying them to Equation 4 Genereux got

\begin{align*}
  u_{p_E} = \sqrt{
    \left( u_{E_{baseflow}} \frac{E_{rain} - E_{streamflow}}
                                 {\left( E_{rain} - E_{baseflow} \right)^2} \right)^2 +
    \left( u_{E_{rain}} \frac{E_{streamflow} - E_{baseflow}}
                             {\left( E_{rain} - E_{baseflow} \right)^2} \right)^2 +
    \left( u_{E_{streamflow}} \frac{-1}
                                   {E_{rain} - E_{baseflow}} \right)^2
  }
\end{align*}

Since we know that, for our analyses, we have identical analytical uncertainty for any measurement of an isotope $E$, such that

\begin{align*}
  u_{E_{streamflow}} = u_{E_{rain}} = u_{E_{baseflow}} = u_E
\end{align*}

With this approach, I get $p_{\delta^{18}O} = `r p$d18O` \pm `r u_p_d18O_Genereux`$ and $p_{\delta^{2}H} = `r p$d2H` \pm `r u_p_d2H_Genereux`$, such that following (2.3) $p = `r p_rain_Genereux` \pm `r u_p_rain_Genereux`$.

**Problem: $u_p$ from (2.2) $\ne$ that from (3.2)**

# References {-}

Genereux (1998) *Water Resources Research* 34(4):915--919
